A finite-state morphological transducer for Kyrgyz

### 2012-03-13 paper/paper.tex
more tweaking

\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
\setromanfont[IPA]{FreeSerif}
%\setromanfont[IPA]{Times New Roman}
%\renewcommand{\sfdefault}{phv}
%\renewcommand{\rmdefault}{ptm}
%\renewcommand{\ttdefault}{pcr}

@@

\begin{table*}[htbp]
	\caption{examples of different affirmative / negative alternations in finite verb forms}\label{irregnegs}
	\centering
	\begin{tabular}{lll}
		\toprule
		tense/aspect & ending $+$ person series & ending $+$ person series \\
		\midrule
		recent eyewitness past & -/DI/ $+$ P4\footnotemark{} & -/GAн жок/ $+$ P3\\
		non-recent past & -/GAн/ $+$ P3 & -/GAн эмес/ $+$ P3\\
		non-recent evidential past & -/GAн экен/ $+$ P3 & -/GAн эмес экен/ $+$ P3\\
		past habitual & -/чU/ $+$ P3 & -/чU эмес/ $+$ P3\\
		recent evidential past & -/(I)птIр/ $+$ P3 & -/BAптIр/ $+$ P3\\
		habitual/future & -/E/ $+$ P6 & -/BAй/ $+$ P6\\
		%FIXME: /й/ here is technically /E/, but it's always realised as [й] after a vowel so will always be [й] here; how should we talk about it here?  Is this actually treated as regular in kymorph?
		\bottomrule
	\end{tabular}
\end{table*}

@@

\begin{table*}[htbp]
	\caption{combinations of possessive suffixes with case suffixes}\label{irregposcase}
	\centering
	\begin{tabular}{llllll}
		\toprule
		case & morphology & 1st person singular & 2nd person sing. & 3rd person & 1st person plural \\
		\midrule
		nom & — & -(I)м & -(I)ң & -(S)I & -(I)бIз \\
		acc & -NI & -(I)мдI & -(I)ңдI & -\textbf{(S)Iн} & -(I)бIздI \\
		gen & -NIн & -(I)мдIн & -(I)ңдIн & -(S)IнIн & -(I)бIздIн \\
		loc & -DA & -(I)мдA & -(I)ңдA & -\textbf{(S)IндA} & -(I)бIздA \\
		abl & -DAн & -(I)мдAн, -\textbf{(I)мAн} & -(I)ндAн, -\textbf{(I)ңAн} & -\textbf{(S)IнAн} & -(I)бIздAн \\
		dat & -GA & -\textbf{(I)мA} & -\textbf{(I)ңA} & -\textbf{(S)IнA} & -(I)бIзгA \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\begin{table}[htbp]
	\centering
	\caption{vowel harmony for archiphoneme \{I\}}\label{vowelharmonyI}
	\subtable{
		\begin{tabular}{cc}
			\toprule
			after & result \\
			\midrule
			и & и \\
			ү & ү \\
			е & и \\
			ө & ү \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{cc}
			\toprule
			after & result \\
			\midrule
			ы & ы \\
			у & у \\
			а & ы \\
			о & у \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

\begin{table}[htbp]
	\centering
	\caption{vowel harmony for archiphoneme \{A\}}\label{vowelharmonyA}
	\subtable{
		\begin{tabular}{ll}
			\toprule
			after & result \\
			\midrule
			и & е \\
			ү & ө \\
			е & е \\
			ө & ө \\
			\bottomrule
		\end{tabular}
	}
	%\qquad
	\hspace{2em}
	\subtable{
		\begin{tabular}{ll}
			\toprule
			after & result \\
			\midrule
			ы & а \\
			у & а \\
			а & а \\
			о & о \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

\begin{table}[htbp]
	\caption{examples comparing voicing and devoicing in Kyrgyz}\label{exvoicing}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-DA/ & [алмада] & \eng{apple--\gmk{LOC}} \\
			/каз-DA/ & [казда] & \eng{goose--\gmk{LOC}} \\
			/баш-DA/ & [башта] & \eng{head--\gmk{LOC}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-GA/ & [алмага] & \eng{apple--\gmk{DAT}} \\
			/каз-GA/ & [казга] & \eng{goose--\gmk{DAT}} \\
			/баш-GA/ & [башка] & \eng{head--\gmk{DAT}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

\begin{table}[htbp]
	\caption{examples comparing desonorisation of \{N\} and \{L\}}\label{exdesonorisation}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-NI/ & [алманы] & \eng{apple--\gmk{ACC}} \\
			/сыр-NI/ & [сырды] & \eng{secret--\gmk{ACC}} \\
			/каз-NI/ & [казды] & \eng{goose--\gmk{ACC}} \\
			/баш-NI/ & [башты] & \eng{head--\gmk{ACC}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-LAр/ & [алмалар] & \eng{apple--\gmk{PL}} \\
			/сыр-LAр/ & [сырлар] & \eng{secret--\gmk{PL}} \\
			/каз-LAр/ & [каздар] & \eng{goose--\gmk{PL}} \\
			/баш-LAр/ & [баштар] & \eng{head--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

Kyrgyz has a number of nouns that underlyingly end with a consonant cluster consisting of two liquids, but which are split by an epenthetic vowel syllable-finally.
\begin{table}[htbp]
	\caption{examples of /рн/-final nouns surfacing with epenthetic vowel}\label{exrnvowel}
	\centering
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/мурн/ & [мурун] & \eng{nose} \\
			/мурн-LAр/ & [мурундар] & \eng{nose--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн/ & [орун] & \eng{place} \\
			/орн-LAр/ & [орундар] & \eng{places--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}
However, when a vowel-initial suffix follows the noun, the epenthetic vowel is absent, and all rules of consonant unfaithfulness involving consonant clusters are enforced[?].
%FIXME: "[?]"
%FIXME: make better references to the tables
\begin{table}[htbp]
	\caption{examples of /рн/-final nouns surfacing with no epenthetic vowel}\label{exrnnovowel}
	%\centering
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			%FIXME: should the glosses here and below be morpheme-by-morpheme for consistency, or is a more colloquial gloss okay?
			/мурн-(S)I/ & [мурду] & \eng{his/her/its nose} \\
			/мурн-Iм-DAн/ & [мурдуман] & \eng{from my nose} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{0.5em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн-(S)I/ & [орду] & \eng{his/her/its place} \\
			/орн-Iм-DAн/ & [ордуман] & \eng{from my place} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

>>>>

\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
\setromanfont[IPA]{FreeSerif}
%\setromanfont[IPA]{Times New Roman}
%\renewcommand{\sfdefault}{phv}
%\renewcommand{\rmdefault}{ptm}
%\renewcommand{\ttdefault}{pcr}
\usepackage[small,bf]{caption}

@@

\begin{table*}[htbp]
	\caption{Examples of different affirmative / negative alternations in finite verb forms}\label{irregnegs}
	\centering
	\begin{tabular}{lll}
		\toprule
		tense/aspect & ending $+$ person series & ending $+$ person series \\
		\midrule
		recent eyewitness past & -/DI/ $+$ P4\footnotemark{} & -/GAн жок/ $+$ P3\\
		non-recent past & -/GAн/ $+$ P3 & -/GAн эмес/ $+$ P3\\
		non-recent evidential past & -/GAн экен/ $+$ P3 & -/GAн эмес экен/ $+$ P3\\
		past habitual & -/чU/ $+$ P3 & -/чU эмес/ $+$ P3\\
		recent evidential past & -/(I)птIр/ $+$ P3 & -/BAптIр/ $+$ P3\\
		habitual/future & -/E/ $+$ P6 & -/BAй/ $+$ P6\\
		%FIXME: /й/ here is technically /E/, but it's always realised as [й] after a vowel so will always be [й] here; how should we talk about it here?  Is this actually treated as regular in kymorph?
		\bottomrule
	\end{tabular}
\end{table*}

@@

\begin{table*}[htbp]
	\caption{Combinations of possessive suffixes with case suffixes}\label{irregposcase}
	\centering
	\begin{tabular}{llllll}
		\toprule
		case & morphology & 1st person singular & 2nd person sing. & 3rd person & 1st person plural \\
		\midrule
		nom & — & -(I)м & -(I)ң & -(S)I & -(I)бIз \\
		acc & -NI & -(I)мдI & -(I)ңдI & -\textbf{(S)Iн} & -(I)бIздI \\
		gen & -NIн & -(I)мдIн & -(I)ңдIн & -(S)IнIн & -(I)бIздIн \\
		loc & -DA & -(I)мдA & -(I)ңдA & -\textbf{(S)IндA} & -(I)бIздA \\
		abl & -DAн & -(I)мдAн, -\textbf{(I)мAн} & -(I)ндAн, -\textbf{(I)ңAн} & -\textbf{(S)IнAн} & -(I)бIздAн \\
		dat & -GA & -\textbf{(I)мA} & -\textbf{(I)ңA} & -\textbf{(S)IнA} & -(I)бIзгA \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\begin{table}[htbp]
	\centering
	\caption{Vowel harmony for archiphoneme \{I\}}\label{vowelharmonyI}
	\subtable{
		\begin{tabular}{cc}
			\toprule
			after & result \\
			\midrule
			и & и \\
			ү & ү \\
			е & и \\
			ө & ү \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{cc}
			\toprule
			after & result \\
			\midrule
			ы & ы \\
			у & у \\
			а & ы \\
			о & у \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

\begin{table}[htbp]
	\centering
	\caption{Vowel harmony for archiphoneme \{A\}}\label{vowelharmonyA}
	\subtable{
		\begin{tabular}{ll}
			\toprule
			after & result \\
			\midrule
			и & е \\
			ү & ө \\
			е & е \\
			ө & ө \\
			\bottomrule
		\end{tabular}
	}
	%\qquad
	\hspace{2em}
	\subtable{
		\begin{tabular}{ll}
			\toprule
			after & result \\
			\midrule
			ы & а \\
			у & а \\
			а & а \\
			о & о \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

\begin{table}[htbp]
	\caption{Examples comparing voicing and devoicing in Kyrgyz}\label{exvoicing}
\begin{center}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-DA/ & [алмада] & \eng{apple--\gmk{LOC}} \\
			/каз-DA/ & [казда] & \eng{goose--\gmk{LOC}} \\
			/баш-DA/ & [башта] & \eng{head--\gmk{LOC}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-GA/ & [алмага] & \eng{apple--\gmk{DAT}} \\
			/каз-GA/ & [казга] & \eng{goose--\gmk{DAT}} \\
			/баш-GA/ & [башка] & \eng{head--\gmk{DAT}} \\
			\bottomrule
		\end{tabular}
	}
\end{center}
\end{table}

@@

\begin{table}[htbp]
	\caption{Examples comparing desonorisation of \{N\} and \{L\}}\label{exdesonorisation}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-NI/ & [алманы] & \eng{apple--\gmk{ACC}} \\
			/сыр-NI/ & [сырды] & \eng{secret--\gmk{ACC}} \\
			/каз-NI/ & [казды] & \eng{goose--\gmk{ACC}} \\
			/баш-NI/ & [башты] & \eng{head--\gmk{ACC}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-LAр/ & [алмалар] & \eng{apple--\gmk{PL}} \\
			/сыр-LAр/ & [сырлар] & \eng{secret--\gmk{PL}} \\
			/каз-LAр/ & [каздар] & \eng{goose--\gmk{PL}} \\
			/баш-LAр/ & [баштар] & \eng{head--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

@@

Kyrgyz has a number of nouns that underlyingly end with a consonant cluster consisting of two liquids, but which are split by an epenthetic vowel syllable-finally.
\begin{table}[htbp]
	\caption{Examples of /рн/-final nouns surfacing with epenthetic vowel}\label{exrnvowel}
	\centering
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/мурн/ & [мурун] & \eng{nose} \\
			/мурн-LAр/ & [мурундар] & \eng{nose--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн/ & [орун] & \eng{place} \\
			/орн-LAр/ & [орундар] & \eng{places--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}
However, when a vowel-initial suffix follows the noun, the epenthetic vowel is absent, and all rules of consonant unfaithfulness involving consonant clusters are enforced[?].
%FIXME: "[?]"
%FIXME: make better references to the tables
\begin{table}[htbp]
	\caption{Examples of /рн/-final nouns surfacing with no epenthetic vowel}\label{exrnnovowel}
	%\centering
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			%FIXME: should the glosses here and below be morpheme-by-morpheme for consistency, or is a more colloquial gloss okay?
			/мурн-(S)I/ & [мурду] & \eng{his/her/its nose} \\
			/мурн-Iм-DAн/ & [мурдуман] & \eng{from my nose} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{0.5em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн-(S)I/ & [орду] & \eng{his/her/its place} \\
			/орн-Iм-DAн/ & [ордуман] & \eng{from my place} \\
			\bottomrule
		\end{tabular}
	}
\end{table}
